Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-08-31 Paper.Rmd
modified data section

## Datasets
<!-- each dataset's description is in suppmat file-->

>>>>

## Data \label{Data}

For bivariate analysis we used four datasets which are briefly described as follows;
to begin with **Soil C and N data**, to explore the correlation of soil organic C and total soil N across a wide geographic range (conterminous US), ecosystem type, and traceable methodology we used the RaCA (Rapid Carbon Assessment)
database [@Wills2014, **or Wills and Loecke pending???**]. Specifically, the soil organic C and total soil N
stocks (Mg C or N m$^{-2}$ ) to 1m of soil depth from the central pedon of the 5930 sites were included in this
study.

Next, the [**BMR data**](https://github.com/sghosh89/BIVAN/blob/master/Data/BMR/bmr-contrasts.csv) contains the trait contrasts from phylogenetically independent contrasts [@SukumaranH2010] of data reported by [@GenoudIM2017] on the basal metabolic rate (BMR) and mass of mammals species. Specifically, the "SELECT" BMR data in their database were used for those species for which data were available. A phylogeny in NEXUS [@NEXUS] format was extracted from the supplementary information of [@GenoudIM2017]. Version 4.4.0 of the DendroPy library [@SukumaranH2010] was used to calculate phylogenetic independent contrasts using Python 2.7. For each contrast, the path length spanned by the contrast (sum of the branch lengths) and the contrast for the mass, natural logarithm of the BMR and natural logarithm of the mass were written as a row in the bmr-contrasts.csv file.

[**Third dataset**](https://github.com/sghosh89/BIVAN/blob/master/Data/Anolis/contrasts.csv) was a raw data on measurements of anatomical features of several species of *Anolis* lizards, which was downloaded
from the Dryad data package [@MahlerIRL2013Data] associated with [@MahlerIRL2013].
The [contrasts.csv file](https://github.com/sghosh89/BIVAN/blob/master/Data/Anolis/contrasts.csv) was produced by using version 4.4.0 of the DendroPy library [@SukumaranH2010]  to calculate phylogenetic independent contrasts on the the data in the GA_Anolis_traits.csv file of Mahler et at. using their maximum clade credibility tree (GA_Anolis_MCC.nex) as the phylogeny.

Absolute values of traits for each contrast are considered to construct corresponding copula for second and third datasets.

Fourth, we used **Ceder-creek data for e120** which resulted from a biodiversity experiment, carried out by Tilman and his co-warkers [@Tilman2001; @Tilman2006] over many years to collect [plant above ground biomass](https://www.cedarcreek.umn.edu/research/data/dataset?ple120) and [plant species percent cover](https://www.cedarcreek.umn.edu/research/data/dataset?pce120) data. These datasets were used to calculate plant-aboveground biomass and Shannon's diversity index (H), averaged over each plot (9m by 9m),
for year 2000.

The raw data and copula for each four datasets are shown along each column of Figure \ref{fig_biv_raw_cop}.

For multivariate analysis, we applied four datasets; first two data sets [for aphid-count and aphid-firstflight date] were extracted from the Rothamsted Insect Survey suction-trap data [@Macaulay1988, @Harrington2014]. These two annual timesries (1976-2010) measurements are from 11 locations acroos UK for 20 aphid species. We used the same data format as used by @Sheppard2016 for their analysis on the impact of climate on aphids' spatial synchrony. In our case, we chose species=10 [Name : "Green spruce aphid"] for aphid-abundance data and species=11 [Name : "Leaf-curling plum aphid"] for aphid-firstflight data and considered only those locations which had atleast 30 years of data (see Figure \ref{SM-fig_aphid_map}). Figure \ref{fig_aphid_raw_cop} represents raw and copula data plot for aphid count ("Green spruce aphid") and aphid-firstflight date ("Leaf-curling plum aphid") taken from two locations.

Our third dataset for multivariate analysis is a plankton-count data collected at 26 locations over the areas of the North Sea and British seas for 22 species and for 56 years (1958-2013). These data were obtained from the Continuous Plankton Recorder (CPR) data set and maintained by the Sir Alistair Hardy Foundation for Ocean Science (SAHFOS). The CPR is a device which is towed across the ocean behind a ship, filtering plankton from seawater with a continuously
unwinding silk ribbon. The ribbon was cut into 4 inch segments, each corresponding to a particular time and
location in the ocean. Organisms were identified and counted (possibly with sub-sampling) microscopically [@Colebrook1979]. All the samples available for a given variable in a given area in a given month of a given year were
averaged, producing a monthly time series for each year. Where data for a given month was unavailable the missing value was replaced with a median value for that month (drawn from all years in which data was available). Monthly values were converted to yearly values by averaging over all months. This approach was used by [@Sheppard2017] earlier. But here we only allowed maximum 4 months in a year with missing data which were replaced by those medians. If a year got 5 or more missing monthly data value then we set NA value for that particular year. We used data for species=16 [Name : "Ceratium furca"] which is a dicoflagellate commonly found in UK seas. Among 26 locations, only those having atleast 45 years time series data, were considered (see Figure \ref{SM-fig_plankton_map}).

Finally, methane ($CH_{4}$) fluxes between the soil or water surface and the troposphere were measured at the
Great Miami Wetland Mitigation Bank ($36^{\circ}46^{'}51^{''}$ N, $84^{\circ}20^{'}26^{''}$ W) in Trotwood, Montgomery County, Ohio. Additional information on the history of the site and soil types is detailed in @Jarecke2016. Briefly, the closed static flux chamber method was used [@Holland1999] to measure $CH_{4}$ fluxes at 28 locations across a hydrologic gradient (upland to seasonal wetland) at
daily to weekly interval from September 2015 to September 2016. **See and Smyth et al, pending for
additional details???** on greenhouse gas flux methods for this site. We threw out all locations that had fewer than 50 dates data.

### 2018-08-31 SupportingInformation.Rmd
modified data section

updated on `r Sys.Date()`

@@

```{r cop_pedag_fig,echo=F,results="hide"}
set.seed(seed)
library(VineCopula)
f<-function(par,destau,desLTmUT,fam){
  cop<-BiCop(family=fam,par=par[1],par2=par[2])
  tau<-cop$tau 
  LT<-cop$taildep$lower
  UT<-cop$taildep$upper
  LTmUT<-LT-UT
  z<-(tau-destau)^2
  z<-z+(LTmUT-desLTmUT)^2
  return(z)
}
#----------------------------------------------------------
pdf("./Results/copula_pedagog_fig_BB1.pdf",height=8,width=20)
op<-par(mfrow=c(2,5),mar=c(3,3,3,3), mgp=c(1.5,0.5,0))

@@

\newpage
# Descriptions of data \label{Description_data}
<!--Each data contributor writes one to a few paragraphs about their data.-->

## Soil C and N data  \label{SoilCN_data}

To explore the correlation of soil organic C and total soil N across a wide geographic range (conterminous
US), ecosystem type, and traceable methodology we used the RaCA (Rapid Carbon Assessment)
database [@Wills2014, **or Wills and Loecke pending???**]. Specifically, the soil organic C and total soil N
stocks (Mg C or N m$^{-2}$ ) to 1m of soil depth from the central pedon of the 5930 sites were included in this
study.

The raw data are shown in Figure \ref{fig_raw_cop_soilCN}(A). The copula is shown in Figure \ref{fig_raw_cop_soilCN}(B).

@@

\begin{figure}[!h]
\textbf{ \hspace{5.5 cm} (A) \hspace{7 cm} (B)  } \\
\begin{center}
\includegraphics[width=3 in]{./Results/SoilCNraw.pdf}
\includegraphics[width=3 in]{./Results/SoilCNcop.pdf}
\caption[\textbf{(A)} Plot of the raw soil C and N data, \textbf{(B)} Copula plot of soil C and N data]{\textbf{(A)} Plot of the raw soil C and N data, \textbf{(B)} Copula plot of soil C and N data. Obtained by applying ranks separately to the C and N data, and then dividing the ranks by $n+1$ where $n = `r nrow(d)` $ is the number of sampling locations (see \nameref{MT-Model_selection} of manuscript).\label{fig_raw_cop_soilCN}}
\end{center}
\end{figure}

## BMR data  \label{BMR_data}

[**bmr-contrasts.csv**](https://github.com/sghosh89/BIVAN/blob/master/Data/BMR/bmr-contrasts.csv) : 
This contains the trait contrasts from phylogenetically independent contrasts [@SukumaranH2010] 
of data reported by [@GenoudIM2017] on the basal metabolic rate (BMR)
and mass of mammals species.
Specifically, the "SELECT" BMR data in their database were used for
 those species for which data were available.
A phylogeny in NEXUS [@NEXUS] format was extracted from the 
 supplementary information of [@GenoudIM2017].
Version 4.4.0 of the DendroPy library [@SukumaranH2010] was used to calculate 
  phylogenetic independent contrasts using Python 2.7.
For each contrast, the path length spanned by the contrast (sum of the branch
  lengths) and the contrast for the mass, natural logarithm of the BMR and natural
  logarithm of the mass were written as a row in the bmr-contrasts.csv file.

```{r read_BMR_data, echo=FALSE, results="markup"}
set.seed(seed)
#---------------------------------------------------------
BMR_dat<-read.csv("./Data/BMR/bmr-contrasts.csv",header = T)
#class(BMR_dat)
knitr::kable(head(BMR_dat[,c(1,4,5)] ,3),format='pandoc',
             caption = "First few rows of raw BMR data.\\label{tab_rawdata_BMR}",
             booktabs=T) # display first few rows
```

Table \ref{tab_rawdata_BMR} displays the raw data for first few contrasts. The raw data are shown in Figure \ref{fig_raw_cop_BMR}(A). The copula [using absolute values of each column in Table \ref{tab_rawdata_BMR}] is shown in Figure \ref{fig_raw_cop_BMR}(B).

@@

\begin{figure}[!h]
\textbf{ \hspace{5.5 cm} (A) \hspace{7 cm} (B)  } \\
\begin{center}
\includegraphics[width=3 in]{./Results/BMR_results/rawdataplot_lnBMR_vs_lnmass.pdf}
\includegraphics[width=3 in]{./Results/BMR_results/copula_lnBMR_vs_lnmass.pdf}
\caption[\textbf{(A)} Plot of the raw ln(BMR) of contrasts vs. ln(mass) of contrasts \textbf{(B)} Copula plot for ln(BMR) vs. ln(mass) contrasts taken as their absolute values]{\textbf{(A)} Plot of the raw ln(BMR) of contrasts vs. ln(mass) of contrasts \textbf{(B)} Copula plot for absolute values of ln(BMR) vs. ln(mass) for all $`r nrow(v_BMR)`$ contrasts. Obtained by applying ranks separately to each variable, and then dividing the ranks by $n+1$ where $n=`r nrow(v_BMR)`$ is the number of contrasts (see \nameref{MT-Model_selection} of manuscript).\label{fig_raw_cop_BMR}}
\end{center}
\end{figure}

## Phylogenetic data : Anolis  \label{Phylo_data}

[**contrasts.csv**](https://github.com/sghosh89/BIVAN/blob/master/Data/Anolis/contrasts.csv) : 
Raw data on measurements of anatomical features of several species of *Anolis* lizards was downloaded
from the Dryad data package [@MahlerIRL2013Data] associated with
[@MahlerIRL2013].
The contrasts.csv file was produced by using version 4.4.0 of the DendroPy 
  library [@SukumaranH2010]  to calculate 
  phylogenetic independent contrasts on the the data in the GA_Anolis_traits.csv
  file of Mahler et at. using their maximum clade credibility tree (GA_Anolis_MCC.nex)
  as the phylogeny.

```{r read_phylo_data, echo=F, results="markup"}
set.seed(seed)
phylo_dat<-read.csv("./Data/Anolis/contrasts.csv",header = T)
#class(phylo_dat) #it's a data.frame
# no NA/missing value in this dataset
tab_phylo<-phylo_dat[,c(1,7,8)] 
knitr::kable(head(tab_phylo,3),format='pandoc',
             caption = "First few rows of raw phylo data.\\label{tab_rawdata_phylo}",
             booktabs=T) # display first few rows
```

Table \ref{tab_rawdata_phylo} displays the raw data for first few contrasts for two selective phylogenetic traits, namely, AVG.ltoe.IV (average length of fourth hindtoe) and AVG.humerus (average length of humerus). The raw data are shown in Figure \ref{fig_raw_cop_AVG_humerus_ltoe}(A). The copula [taking absolute values of each column in Table \ref{tab_rawdata_phylo}] is shown in Figure \ref{fig_raw_cop_AVG_humerus_ltoe}(B).

@@

\begin{figure}[!h]
\textbf{ \hspace{5.5 cm} (A) \hspace{7 cm} (B)  } \\
\begin{center}
\includegraphics[width=3 in]{./Results/Anolis_results/contrasts_csv_results/rawdataplot_AVG_ltoe_humerus.pdf}
\includegraphics[width=3 in]{./Results/Anolis_results/contrasts_csv_results/copula_AVG_ltoe_humerus.pdf}
\caption[\textbf{(A)} Plot of the contrasts for AVG.ltoe.IV vs. AVG.humerus phylogenetic traits. \textbf{(B)} Copula plot of contrasts for AVG.ltoe.IV vs. contrasts for AVG.humerus phylogenetic traits(taken as absolute values)]{\textbf{(A)} Plot of the contrasts for AVG.ltoe.IV vs. contrasts for AVG.humerus phylogenetic traits. \textbf{(B)} Copula plot of absolute values of contrasts for AVG.ltoe.IV vs. contrasts for AVG.humerus phylogenetic traits. Obtained by applying ranks separately to each variable, and then dividing the ranks by $n+1$ where $n=`r nrow(v_contrasts_67)`$ is the number of contrasts (see \nameref{MT-Model_selection} of manuscript). \label{fig_raw_cop_AVG_humerus_ltoe}}
\end{center}
\end{figure}

## Ceder-creek data for e120 \label{ceder-creek_data}

These biodiversity experiment data sets, for [plant above ground biomass](https://www.cedarcreek.umn.edu/research/data/dataset?ple120) and for [plant species percent cover](https://www.cedarcreek.umn.edu/research/data/dataset?pce120), are collected by Tilman et al. over many years. We used thsese datasets [@Tilman2001; @Tilman2006] to calculate plant-aboveground biomass and Shannon's diversity index (H), averaged over each plot (9m by 9m),
for year 2000.

@@

\begin{figure}[!h]
\textbf{ \hspace{5.5 cm} (A) \hspace{7 cm} (B)  } \\
\vspace{-0.2 cm}
\begin{center}
\includegraphics[width=3 in]{./Results/Ceder_creek_results/raw_HB_2000.pdf}
\includegraphics[width=3 in]{./Results/Ceder_creek_results/cop_HB_2000.pdf}
\caption[\textbf{(A)} Plot of the plant-aboveground avg. Biomass vs. avg. Shannon's index (H). \textbf{(B)} Copula plot of plant-aboveground avg. Biomass vs. avg. Shannon's index.]{\textbf{(A)} Plot of the plant-aboveground avg. Biomass vs. avg. Shannon's index (H). \textbf{(B)} Copula plot of plant-aboveground avg. Biomass vs. avg. Shannon's index. Obtained by applying ranks separately to each variable, and then dividing the ranks by $n+1$ where $n=`r nrow(cc_cop_2000)`$ is the number of (9m by 9m)plots. (see \nameref{MT-Model_selection} of manuscript). \label{fig_raw_cop_cc2000}}
\end{center}
\end{figure}

## Abundance data for the green spruce aphid and first flight date data for the leaf-curling plum aphid in the UK\label{Aphid_data}
   * Aphid data contains 3 measurements [count, firstflight, flight-duration] at 11 locations for 20 species and for 35 years.
   * We threw out all locations that had fewer than 30 years data. 
   * We use species=10 [Name : "Green spruce aphid"] for aphid-count data. 
   * We use species=11 [Name : "Leaf-curling plum aphid"] for aphid-firstflight data.

## Abundance data for *Ceratium furca* in UK seas \label{Pns_data}
   * Plankton North sea dataset collects data at 26 locations for 22 species and for 56 years.
   * We threw out all locations that had fewer than 45 years data.
   * We use species=16 [Name : "Ceratium furca"] for Plankton_North_Sea data.
   * *Ceratium furca* is a dicoflagellate commonly found in UK seas.
   * note, when we edit this, we cannot call it the North Seas since data come also from other seas around the UK.

## Methane-flux data  \label{Methane_data}

@@

* We threw out all locations that had fewer than 50 dates data.

>>>>

updated on `r Sys.Date()`
\pagebreak

@@

```{r cop_pedag_fig,echo=F,results="hide"}
set.seed(seed)
library(VineCopula)
f<-function(par,destau,desLTmUT,fam){
  cop<-BiCop(family=fam,par=par[1],par2=par[2])
  tau<-cop$tau 
  LT<-cop$taildep$lower
  UT<-cop$taildep$upper
  LTmUT<-LT-UT
  z<-(tau-destau)^2
  z<-z+(LTmUT-desLTmUT)^2
  return(z)
}

pdf("./Results/copula_pedagog_fig_BB1.pdf",height=8,width=20)
op<-par(mfrow=c(2,5),mar=c(3,3,3,3), mgp=c(1.5,0.5,0))

@@

@@

```{r read_BMR_data, echo=FALSE, results="markup"}
set.seed(seed)
#---------------------------------------------------------
BMR_dat<-read.csv("./Data/BMR/bmr-contrasts.csv",header = T)
#class(BMR_dat)
#knitr::kable(head(BMR_dat[,c(1,4,5)] ,3),format='pandoc',
#             caption = "First few rows of raw BMR data.\\label{tab_rawdata_BMR}",
#             booktabs=T) # display first few rows
```

@@

```{r read_phylo_data, echo=F, results="markup"}
set.seed(seed)
phylo_dat<-read.csv("./Data/Anolis/contrasts.csv",header = T)
#class(phylo_dat) #it's a data.frame : no NA/missing value in this dataset
#tab_phylo<-phylo_dat[,c(1,7,8)] 
#knitr::kable(head(tab_phylo,3),format='pandoc',
#             caption = "First few rows of raw phylo data.\\label{tab_rawdata_phylo}",
#             booktabs=T) # display first few rows
```

@@

@@

```{r map_plot_aphid_plankton,results="hide",echo=F}
source("aphid_data_calling.R")
source("good_loclist.R")
library(maps)

#----------------- Aphid map--------------------------------------------
d0_count<-as.matrix(read.csv("./Data/Aphid_data/APHID_DATA_RAW/aphidtscount141117.csv",header=F))
data_aphid_count<-d_allsp_data(d0_count)

d0_firstflight<-as.matrix(read.csv("./Data/Aphid_data/APHID_DATA_RAW/aphidtsfirstflight141117.csv",header=F))
data_aphid_ff<-d_allsp_data(d0_firstflight)

selective_loc1<-good_loclist(d_allsp=data_aphid_count,sp=10,data_pt_thrs=30)
selective_loc2<-good_loclist(d_allsp=data_aphid_ff,sp=11,data_pt_thrs=30)
all(selective_loc1==selective_loc2)==T

longs<-c(-4.567,0.57,-3.069,-3.312,-2.637,-1.682,-2.763,-0.356,-3.454,0.427,0.939)
lats<-c(55.477,52.26,56.457,55.949,52.125,55.213,53.854,51.807,50.628,51.733,51.185)

locno<-length(longs)
df_loc_info_org<-data.frame(loc_no=c(1:length(lats)),longs=longs,lats=lats)
df_loc_info<-df_loc_info_org[selective_loc1,]
xlim=c(-10,2)
ylim=c(50, 60)

pdf("./Results/aphid_map_plot.pdf")
map("world", fill=TRUE, col="grey84", bg="white",border=0, xlim=xlim, ylim=ylim)
map.axes(cex.axis=2.5)
points(x = df_loc_info_org$longs, y = df_loc_info_org$lats, col="red",pch=16)
points(x = df_loc_info$longs, y = df_loc_info$lats, col="blue", pch=0, cex=1.5)
text(df_loc_info_org$loc_no, x = df_loc_info_org$longs, y = df_loc_info_org$lats, col="red",  pos=1)
dev.off()

#--------------Plankton map----------------------------

source("plankton_north_sea_data_calling.R")
d0_plankton_north_sea<-as.matrix(read.csv("./Data/Plankton_North_Sea_data/planktontimeseries201117_4.csv",header=F))
data_plankton_north_sea<-d_allsp_data_plankton_north_sea(d0_plankton_north_sea)

longs<-read.csv("./Data/Plankton_North_Sea_data/boxcornerlongs201117.csv",header=F)
lats<-read.csv("./Data/Plankton_North_Sea_data/boxcornerlats201117.csv",header=F)
longs<-longs[[1]]
lats<-lats[[1]]

df_loc_info_org<-data.frame(loc_no=c(1:length(lats)),longs=longs,lats=lats)
selective_loc<-good_loclist(d_allsp=data_plankton_north_sea,sp=16,data_pt_thrs=45)
df_loc_info<-df_loc_info_org[selective_loc,]
xlim=c(-12,10)
ylim=c(48, 60)

pdf("./Results/plankton_map_plot.pdf")
map("world", fill=TRUE, col="grey84", bg="white",border=0, xlim=xlim, ylim=ylim)
map.axes(cex.axis=2.5)
abline(h=seq(from=ylim[1],to=ylim[2],by=2), v=seq(from=xlim[1],to=xlim[2],by=2), col="black", lty=2)
points(x = df_loc_info_org$longs+1, y = df_loc_info_org$lats+1, col="red",pch=16)
points(x = df_loc_info$longs+1, y = df_loc_info$lats+1, col="blue", pch=0, cex=1.5)
text(df_loc_info_org$loc_no, x = df_loc_info_org$longs+1, y = df_loc_info_org$lats+1, col="red",  pos=1)
dev.off()
```

\begin{figure}[!h]
\begin{center}
\includegraphics[width=8 cm,height=8 cm]{./Results/aphid_map_plot.pdf} 
\caption{Locations of Rothamsted Insect Survey suction-trap across UK are indicated by the red circles 
out of which only blue boxes are considered for our study based on having atleast 30 annual data points. \label{fig_aphid_map}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=8 cm,height=8 cm]{./Results/plankton_map_plot.pdf} 
\caption{CPR data collected from 26 locations ($2^{\circ}$ $\times$ $2^{\circ}$ on map) around UK sea (indicated by the red circles) but out of which only blue boxes are considered for our study based on having atleast 45 annual data points. \label{fig_plankton_map}}
\end{center}
\end{figure}

<!--
## Methane-flux data  \label{Methane_data}

@@

* We threw out all locations that had fewer than 50 dates data.-->
